---
title: "Codes epifitter manuscript"
author: "Kaique Alves & Emerson Del Ponte"
date: "22/09/2020"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(epifitter)
library(tidyverse)
library(cowplot)
```


## Simulating DPCs for different models
### The exponential model
```{r}
exp_model <- sim_exponential(
  N = 100,
  y0 = 0.01,
  dt = 5,
  r = 0.045,
  alpha = 0.2,
  n = 7
)
head(exp_model)
```

The function returns a `data.frame` object with four columns. `replicates` indicate each respective replicate, in `time` we have the time steps, `y` contains the simulated values of disease intensity, and `random_y` contains values of disease intensity after inserting a variation. 

Use the [`ggplot2`](https://ggplot2.tidyverse.org/) package to build impressive graphics!


```{r}
exp_plot = exp_model %>%
  ggplot(aes(time, y)) +
  geom_jitter(aes(time, random_y), size = 3,color = "gray", width = .1) +
  geom_line(size = 1) +
  theme_minimal_hgrid() +
  ylim(0,1)+
  labs(
    title = "Exponential",
    y = "Disease intensity",
    x = "Time"
  )
exp_plot
```

### The Monomolecular model

For the other models is exactly the same pipeline.

```{r}
mono_model <- sim_monomolecular(
  N = 100,
  y0 = 0.01,
  dt = 5,
  r = 0.05,
  alpha = 0.2,
  n = 7
)
head(mono_model)
```

```{r}
mono_plot = mono_model %>%
  ggplot(aes(time, y)) +
  geom_jitter(aes(time, random_y), size = 3, color = "gray", width = .1) +
  geom_line(size = 1) +
  theme_minimal_hgrid() +
  labs(
    title = "Monomolecular",
    y = "Disease intensity",
    x = "Time"
  )
mono_plot
```

### The Logistic model


```{r}
logist_model <- sim_logistic(
  N = 100,
  y0 = 0.01,
  dt = 5,
  r = 0.1,
  alpha = 0.2,
  n = 7
)
head(logist_model)
```

```{r}
logist_plot = logist_model %>%
  ggplot(aes(time, y)) +
  geom_jitter(aes(time, random_y), size = 3,color = "gray", width = .1) +
  geom_line(size = 1) +
  theme_minimal_hgrid() +
  labs(
    title = "Logistic",
    y = "Disease intensity",
    x = "Time"
  )
logist_plot
```

### The Gompertz model


```{r}
gomp_model <- sim_gompertz(
  N = 100,
  y0 = 0.01,
  dt = 5,
  r = 0.07,
  alpha = 0.2,
  n = 7
)
head(gomp_model)
```

```{r}
gomp_plot = gomp_model %>%
  ggplot(aes(time, y)) +
  geom_jitter(aes(time, random_y), size = 3,color = "gray", width = .1) +
  geom_line(size = 1) +
  theme_minimal_hgrid() +
  labs(
    title = "Gompertz",
    y = "Disease intensity",
    x = "Time"
  )
gomp_plot
```

## Combo

Use the function `plot_grid()` from the [`cowplot`](https://wilkelab.org/cowplot/index.html) package to gather all plots into a grid

```{r fig.height=6, fig.width=8}
plot_grid(exp_plot,
          mono_plot,
          logist_plot,
          gomp_plot)
ggsave("figure1.png", dpi = 600, height = 6, width = 9)
```



