---
title: "Uncertainty bands for fitted disease progress curves"
output:
  rmarkdown::html_vignette:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Uncertainty bands for fitted disease progress curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.8
)
```

```{r setup}
library(epifitter)
```

## Why add intervals?

Fitted disease progress curves are usually shown as a single line, but the
line is only an estimate. Confidence bands help communicate the uncertainty
around the fitted epidemic curve.

This article focuses on uncertainty in the fitted mean curve. The bands shown
by `plot_fit()` are not prediction intervals for future observations, and they
do not replace a full model-based analysis of treatment effects. They are a
visual tool for asking how stable the fitted curve is under residual
resampling.

In `epifitter`, confidence bands are added directly in `plot_fit()`:

```r
plot_fit(fit, conf_int = TRUE, ci_method = "bootstrap", nsim = 500)
```

The argument `conf_int` controls whether the interval is drawn. The argument
`ci_method` controls how the interval is estimated.

Use these intervals to compare fitted shapes cautiously. If the scientific
question is a formal comparison among treatments, use a statistical model that
matches the experimental design after defining the target quantity.

## Example data

We first simulate one logistic epidemic with moderate observational noise and
fit the four linearized epidemic models. The noise is intentionally visible in
this example so the confidence bands are easy to inspect.

```{r}
set.seed(123)

epi <- sim_logistic(
  N = 30,
  y0 = 0.01,
  dt = 5,
  r = 0.25,
  alpha = 0.40,
  n = 1
)

fit <- fit_lin(time = epi$time, y = epi$random_y)

head(fit$data)
```

## Residual bootstrap confidence bands

The default bootstrap method resamples residuals from the fitted curve, creates
new plausible disease progress curves at the observed assessment times, refits
the model many times, and builds the band from the distribution of fitted
curves.

This approach assumes that the residual pattern is a reasonable basis for
generating plausible alternative curves. It is useful for visualization, but it
should still be checked against the biology of the system and the assessment
design.

```{r fig.alt="Fitted logistic disease progress curve with a bootstrap confidence band."}
plot_fit(
  fit,
  models = "Logistic",
  conf_int = TRUE,
  ci_method = "bootstrap",
  nsim = 100,
  seed = 123
)
```

For final analyses, increase `nsim`, for example:

```r
plot_fit(fit, conf_int = TRUE, ci_method = "bootstrap", nsim = 500)
```

Larger values of `nsim` make the interval more stable, but also make the plot
slower to compute.

## Wild bootstrap confidence bands

The wild bootstrap method randomly changes the sign of resampled residuals
before refitting the model. This keeps the original assessment times intact and
is available with `ci_method = "wild"`.

Wild bootstrap intervals are often useful as a sensitivity check when residual
variation changes across the range of fitted values.

```{r fig.alt="Fitted logistic disease progress curve with a wild bootstrap confidence band."}
plot_fit(
  fit,
  models = "Logistic",
  conf_int = TRUE,
  ci_method = "wild",
  nsim = 100,
  seed = 123
)
```

## Compare multiple models

Confidence bands can also be drawn for more than one model at a time.

```{r fig.alt="Fitted exponential, logistic, and Gompertz disease progress curves with bootstrap confidence bands."}
plot_fit(
  fit,
  models = c("Exponential", "Logistic", "Gompertz"),
  conf_int = TRUE,
  ci_method = "bootstrap",
  nsim = 100,
  seed = 123
)
```

## Nonlinear fits

The same interface works with nonlinear fits returned by `fit_nlin()` and
`fit_nlin2()`.

```{r fig.alt="Nonlinear fitted logistic disease progress curve with a bootstrap confidence band."}
nlin_fit <- fit_nlin(
  time = epi$time,
  y = epi$random_y,
  starting_par = list(y0 = 0.01, r = 0.1),
  maxiter = 200
)

plot_fit(
  nlin_fit,
  models = "Logistic",
  conf_int = TRUE,
  ci_method = "bootstrap",
  nsim = 100,
  seed = 123
)
```

## Choosing a method

Use `ci_method = "bootstrap"` for the main fitted-curve intervals. Use
`ci_method = "wild"` as a sensitivity check when you want a second residual
bootstrap strategy that preserves the original assessment times.

In very smooth datasets, the confidence band may be narrow. Wider bands usually
appear when the observations are noisier, when there are fewer assessment
times, or when the fitted model is less certain.

In both cases, `level` controls the confidence level:

```r
plot_fit(fit, conf_int = TRUE, ci_method = "bootstrap", level = 0.90)
plot_fit(fit, conf_int = TRUE, ci_method = "bootstrap", level = 0.95)
```

The bands shown by `plot_fit()` are confidence bands for the fitted curve, not
prediction intervals for future observations.

Report `nsim`, `level`, the fitting method, and the interval method when using
these plots in reports or manuscripts.

By default, `plot_fit()` keeps fitted curves and bands on the disease
proportion scale from 0 to 1. To inspect unconstrained fitted values, use:

```r
plot_fit(fit, conf_int = TRUE, y_bounds = NULL)
```
