---
title: "Fitting and selecting models"
author: "Kaique S Alves"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting and selecting models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

## Base functions

`epifitter` provides functions to fit classic two-parameter population dynamics models to disease progress curve (DPC) data: exponential, monomolecular, logistic, and Gompertz.

The goal of fitting these models to DPCs is to estimate epidemiologically meaningful parameters: `y0`, which represents primary inoculum, and `r`, which represents the apparent infection rate. Choosing the model that best describes the epidemic helps users interpret disease progress and compare epidemics more effectively.

Two approaches can be used to obtain the parameters:

-   **Linear regression** models fitted to the transformed disease data required by each model.
-   **Non-linear regression** models fitted to the original disease intensity data.

Both approaches are available in *epifitter*. The simplest way to fit these models to a single epidemic is with `fit_lin()` and `fit_nlin()`. The alternative `fit_nlin2()` allows estimation of a third parameter, the upper asymptote, when the maximum disease intensity is not close to 100%. The `fit_multi()` function extends these workflows to multiple epidemic datasets.

First, we need to load the packages we'll need for this tutorial.

```{r message=FALSE, warning=FALSE}
library(epifitter)
library(ggplot2)
library(dplyr)
library(magrittr)
library(cowplot)
theme_set(cowplot::theme_half_open(font_size = 12))
```

## Basic usage

### Dataset

To use *epifitter*, at least two variables are needed: one representing assessment time and another representing disease intensity as a proportion (for example, incidence, severity, or prevalence). In designed experiments with replicates, a third variable identifying the experimental units is also needed.

Let's simulate a DPC dataset for one epidemic measured in replicated plots. The simulated data resemble a polycyclic epidemic with a sigmoid shape. We can generate this dataset with `sim_logistic()`.

```{r}
dpcL <- sim_logistic(
  N = 100, # duration of the epidemics in days
  y0 = 0.01, # disease intensity at time zero
  dt = 10, # interval between assessments
  r = 0.1, # apparent infection rate
  alpha = 0.2, # level of noise
  n = 7 # number of replicates
)
```

Let's give a look at the simulated dataset.

```{r}
knitr::kable(head(dpcL), digits = 4)
```

The object generated by `sim_logistic()` is a data frame with four columns. The `y` variable contains disease intensity values on a proportional scale (0 \< y \< 1). To facilitate visualization, we can plot the epidemic with the *ggplot2* package.

```{r fig.alt="Plot of simulated logistic epidemics with replicate observations and the underlying mean disease progress curve over time."}
ggplot(
  dpcL,
  aes(time, y,
    group = replicates
  )
) +
  geom_point(aes(time, random_y), shape = 1) + # plot the replicate values
  geom_point(color = "steelblue", size = 2) +
  geom_line(color = "steelblue") +
  labs(
    title = "Simulated 'complete' epidemics of sigmoid shape",
    subtitle = "Produced using sim_logistic()"
  )+
  theme_minimal_hgrid()
```

### Linear regression using `fit_lin()`

The `fit_lin()` requires at least the `time` and `y` arguments. In the example, we will call the `random_y` which represents the replicates. A quick way to call these variables attached to the dataframe is shown below.

```{r}
f_lin <- fit_lin(
  time = dpcL$time,  
  y = dpcL$random_y
)

```

`fit_lin()` outputs a list object which contains several elements. Three elements of the list are shown by default: stats of model fit, Infection rate and Initial Inoculum

```{r}
f_lin
```

#### Model fit stats

The `Stats` element of the list shows how each of the four models predicted the observations based on three measures:

-   Lin's concordance correlation coefficient `CCC` (Lin 2000), a measure of agreement that takes both bias and precision into account
-   Coefficient of determination `r_squared` (R<sup>2</sup>), a measure of precision
-   Residual standard deviation `RSE` for each model.

The four models are sorted from the high to the low `CCC`. As expected because the `sim_logistic` function was used to create the synthetic epidemic data, the the *Logistic* model was superior to the others.

#### Model coefficients

The estimates, and respective standard error and upper and lower 95% confidence interval, for the two coefficients of interest are shown in the `Infection rate` and `Initial inoculum` elements. For the latter, both the back-transformed (estimate) and the linearized estimate are shown.

#### Global stats

The element `f_lin$stats_all` provides a wide format dataframe with all the stats for each model.

```{r}
knitr::kable(f_lin$stats_all, digits = 4)
```

#### Model predictions

The predicted values are stored as a dataframe in the `data` element called using the same `$` operator as above. Both the observed and (`y`) and the back-transformed predictions (`predicted`) are shown for each model. The linearized value and the residual are also shown.

```{r}
knitr::kable(head(f_lin$data), digits = 4)
```

#### Plot of predictions

The `plot_fit()` produces, by default, a panel of plots depicting the observed and predicted values by all fitted models. The arguments `pont_size` and `line_size` that control for the size of the dots for the observation and the size of the fitted line, respectively.

```{r fig.alt="Faceted comparison of observed data and fitted disease progress curves for the candidate linearized models."}
plot_lin <- plot_fit(f_lin,
  point_size = 2,
  line_size = 1
) 

# Default plots
plot_lin 
```

#### Publication-ready plots

The plots are *ggplot2* objects which can be easily customized by adding new layers that override plot paramaters as shown below. The argument `models` allows to select the models(s) to be shown on the plot. The next plot was customized for the logistic model.

```{r fig.alt="Customized plot of the logistic fitted curve with observed disease intensity points over time."}
# Customized plots

plot_fit(f_lin,
  point_size = 2,
  line_size = 1,
  models = "Logistic")+
  theme_minimal_hgrid(font_size =18) +
  scale_x_continuous(limits = c(0,100))+
  scale_color_grey()+
   theme(legend.position = "none")+
  labs(
    x = "Time",
    y = "Proportion disease "
    
  )
```

### Non-linear regression

#### Two-parameters

The `fit_nlin()` function uses the Levenberg-Marquardt algorithm for least-squares estimation of nonlinear parameters. In addition to time and disease intensity, starting values for `y0` and `r` should be given in the `starting_par` argument. The output format and interpretation is analogous to the `fit_lin()`.

> NOTE: If you encounter error messages saying "matrix at initial parameter estimates", try to modify the starting values for the parameters to solve the problem.

```{r message=FALSE, warning=FALSE}
f_nlin <- fit_nlin(
  time = dpcL$time,
  y = dpcL$random_y,
  starting_par = list(y0 = 0.01, r = 0.03)
)

f_nlin
```

We can check the results using `plot_fit`.

```{r fig.alt="Faceted comparison of observed data and fitted curves from nonlinear two-parameter model fits."}
plot_fit(f_nlin) +
  theme_minimal_hgrid()#changing plot theme
```

### Estimating `K` (maximum disease)

In many epidemics the last measure (final time) of a DPC does not reach the maximum intensity and, for this reason, estimation of maximum asymptote (carrying capacity `K`) may be necessary. The `fin_lin2()` provides an estimation of `K` in addition to the estimates provided by `fit_lin()`.

Before demonstrating the function, we can transform our simulated data by creating another variable with `y_random2` with maximum about 0.8 (80%). Simplest way is to multiply the `y_random` by 0.8.

```{r}

dpcL2 = dpcL %>% 
  mutate(random_y = random_y * 0.8)
```

Then we run the `fit_nlin2()` for the new dataset.

```{r message=FALSE, warning=FALSE, fig.alt="Faceted comparison of observed data and fitted curves from nonlinear model fits that estimate the upper asymptote K."}
f_nlin2 <- fit_nlin2(
  time = dpcL2$time,
  y = dpcL2$random_y,
  starting_par = list(y0 = 0.01, r = 0.2, K =  0.6)
)
f_nlin2
plot_fit(f_nlin2)
```

> NOTE: The exponential model is not included because it doesn't have a maximum asymptote. The estimated value of `K` is the expected 0.8.

### Fit models to multiple DPCs

Most commonly, there are more than one epidemics to analyse either from observational or experimental studies. When the goal is to fit a common model to all curves, the `fit_multi()` function is in hand. Each DPC needs an unique identified to further combined in a single data frame.

#### Data

Let's use the `sim_` family of functions to create three epidemics and store the data in a single `data.frame`. The Gompertz model was used to simulate these data. Note that we allowed to the `y0` and `r` parameter to differ the DPCs. We should combine the three DPCs using the `bind_rows()` function and name the identifier (`.id`), automatically created as a character vector, for each epidemics as 'DPC'.

```{r}
epi1 <- sim_gompertz(N = 60, y0 = 0.001, dt = 5, r = 0.1, alpha = 0.4, n = 4)
epi2 <- sim_gompertz(N = 60, y0 = 0.001, dt = 5, r = 0.12, alpha = 0.4, n = 4)
epi3 <- sim_gompertz(N = 60, y0 = 0.003, dt = 5, r = 0.14, alpha = 0.4, n = 4)

multi_epidemic <- bind_rows(epi1,
  epi2,
  epi3,
  .id = "DPC"
)
knitr::kable(head(multi_epidemic), digits = 4)
```

We can visualize the three DPCs in a same plot

```{r fig.alt="Line plot comparing three simulated Gompertz disease progress curves identified by DPC."}

p_multi <- ggplot(multi_epidemic,
       aes(time, y, shape = DPC, group = DPC))+
  geom_point(size =2)+
  geom_line()+
  theme_minimal_grid(font_size =18) +
   labs(
    x = "Time",
    y = "Proportion disease "
    
  )
p_multi
```

Or use `facet_wrap()` for ploting them separately.

```{r fig.height=10, fig.width=6, fig.alt="Faceted display of the three simulated disease progress curves, one panel for each DPC."}
p_multi +
  facet_wrap(~ DPC, ncol = 1)
```

#### Using `fit_multi()`

`fit_multi()` requires at least four arguments: time, disease intensity (as proportion), data and the curve identifier (`strata_cols`). The latter argument accepts one or more strata include as `c("strata1",strata2")`. In the example below, the stratum name is `DPC`, the name of the variable.

By default, the linear regression is fitted to data but adding another argument `nlin = T`, the non linear regressions is fitted instead.

```{r}
multi_fit <- fit_multi(
  time_col = "time",
  intensity_col = "random_y",
  data = multi_epidemic,
  strata_cols = "DPC"
)
```

All parameters of the list can be returned using the \$ operator as below.

```{r}
knitr::kable(head(multi_fit$Parameters), digits = 4)
```

Similarly, all data can be returned.

```{r}
knitr::kable(head(multi_fit$Data), digits = 4)
```

If nonlinear regression is preferred, the `nlim` argument should be set to `TRUE`

```{r}
multi_fit2 <- fit_multi(
  time_col = "time",
  intensity_col = "random_y",
  data = multi_epidemic,
  strata_cols = "DPC",
  nlin = TRUE)
knitr::kable(head(multi_fit2$Parameters), digits = 4)
```

#### Want to estimate K?

If you want to estimate `K`, set `nlin = TRUE` and `estimate_K = TRUE`.

> NOTE: If you do not set both arguments `TRUE`, `K` will not be estimated, because `nlin` defaut is `FALSE`. Also remember that when estimating K, we don't fit the *Exponential* model.

```{r}
multi_fit_K <- fit_multi(
  time_col = "time",
  intensity_col = "random_y",
  data = multi_epidemic,
  strata_cols = "DPC",
  nlin = T,
  estimate_K = T
)
```

```{r}
knitr::kable(head(multi_fit_K$Parameters), digits = 4)
```

### Graphical outputs

Use [`ggplot2`](https://ggplot2.tidyverse.org/) to produce elegant data visualizations of models curves and the estimated parameters.

#### DPCs and fitted curves

The original data and the predicted values by each model are stored in `multi_fit$Data`. A nice plot can be produced as follows:

```{r fig.alt="Facet grid of observed data and fitted disease progress curves for each DPC and model combination."}
multi_fit$Data %>%
  ggplot(aes(time, predicted, color = DPC)) +
  geom_point(aes(time, y), color = "gray") +
  geom_line(size = 1) +
  facet_grid(DPC ~ model, scales = "free_y") +
  theme_minimal_hgrid()+
  coord_cartesian(ylim = c(0, 1))
```

Using the *dplyr* function `filter` only the model of interest can be chosen for plotting.

```{r fig.alt="Plot of observed data and fitted Gompertz curves for each DPC over time."}
multi_fit$Data %>%
  filter(model == "Gompertz") %>%
  ggplot(aes(time, predicted, color = DPC)) +
  geom_point(aes(time, y),
    color = "gray",
    size = 2
  ) +
  geom_line(size = 1.2) +
  theme_minimal_hgrid() +
  labs(
    x = "Time",
    y = "Disease Intensity"
  )
```

#### Apparent infection rate

The `multi_fit$Parameters` element is where all stats and parameters as stored. Let's plot the estimates of the apparent infection rate.

```{r fig.alt="Point and error-bar plot of apparent infection rate estimates with confidence intervals for each DPC."}
multi_fit$Parameters %>%
  filter(model == "Gompertz") %>%
  ggplot(aes(DPC, r)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = r_ci_lwr, ymax = r_ci_upr),
    width = 0,
    size = 1
  ) +
  labs(
    x = "Time",
    y = "Apparent infection rate"
  ) +
  theme_minimal_hgrid()
```

## References

Lin L (2000). A note on the concordance correlation coefficient. Biometrics 56: 324 - 325.
