---
title: "Start here: your first disease progress workflow"
author: "Kaique S. Alves"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Start here: your first disease progress workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: inline
---

```{r setup, message=FALSE, warning=FALSE}
library(epifitter)
library(dplyr)
library(ggplot2)
library(cowplot)
theme_set(cowplot::theme_half_open(font_size = 12))
```

## What epifitter does

`epifitter` helps analyze plant disease progress curves (DPCs): repeated
measurements of disease intensity through time. These curves are common in
plant pathology because they connect epidemic development, host-pathogen
biology, management effects, and quantitative summaries such as apparent
infection rate and area under the disease progress curve.

This article is the shortest path to a complete workflow. You will:

- simulate a synthetic epidemic;
- fit canonical disease progress models;
- inspect fitted parameters and model rankings;
- plot observed and fitted disease progress curves;
- see how the same workflow extends to multiple epidemics.

The examples use simulated data so that the workflow is reproducible, but the
same functions can be applied to observed disease incidence or severity data on
a proportional scale.

## A first epidemic

We start by simulating a logistic epidemic. The column `y` is the underlying
curve and `random_y` represents replicated observations with sampling or
measurement noise.

```{r}
set.seed(1)

epi <- sim_logistic(
  N = 80,
  y0 = 0.01,
  dt = 10,
  r = 0.12,
  alpha = 0.2,
  n = 5
)

knitr::kable(head(epi), digits = 4)
```

```{r fig.alt="Plot of a simulated epidemic showing replicate observations and the underlying disease progress curve over time."}
ggplot(epi, aes(time, y, group = replicates)) +
  geom_point(aes(y = random_y), shape = 1, color = "#8597a4") +
  geom_line(color = "#15616d", linewidth = 0.8) +
  labs(
    title = "Simulated epidemic",
    x = "Time",
    y = "Disease intensity"
  )
```

## Fit candidate models

`fit_lin()` compares the four canonical disease progress models by fitting
linearized forms of each model. This is useful for quick screening and teaching,
but the ranking should be interpreted together with the plot and biological
plausibility.

```{r}
fit <- fit_lin(time = epi$time, y = epi$random_y)
fit
```

`fit$stats_all` contains the full ranking of candidate models.

```{r}
knitr::kable(fit$stats_all, digits = 4)
```

## Visualize predictions

The fitted object stores observed values, transformed values, predictions, and
parameter summaries. `plot_fit()` turns that object into a faceted `ggplot2`
figure.

```{r fig.alt="Faceted plot comparing observed disease intensity values with fitted curves from candidate models."}
plot_fit(fit, point_size = 1.8, line_size = 0.9)
```

## Work with multiple epidemics

Many experiments include multiple treatments, cultivars, blocks, or locations.
Use `fit_multi()` when the same fitting workflow should be applied separately
to each group.

```{r}
epi_a <- sim_gompertz(N = 50, y0 = 0.002, dt = 5, r = 0.10, alpha = 0.2, n = 3)
epi_b <- sim_gompertz(N = 50, y0 = 0.002, dt = 5, r = 0.14, alpha = 0.2, n = 3)

multi_epi <- bind_rows(epi_a, epi_b, .id = "epidemic")

multi_fit <- fit_multi(
  time_col = "time",
  intensity_col = "random_y",
  data = multi_epi,
  strata_cols = "epidemic"
)

knitr::kable(head(multi_fit$Parameters), digits = 4)
```

## Next steps

- Use the PowderyMildew article for a real-data workflow based on the bundled experimental dataset.
- Use the model fitting article for a deeper walkthrough of `fit_lin()`, `fit_nlin()`, `fit_nlin2()`, weighted nonlinear fits, and `fit_multi()`.
- Use the confidence interval article when you need uncertainty bands around fitted curves.
- Use the area-under-the-curve article when your goal is to summarize each epidemic as AUDPC or AUDPS.
- Use the simulation workflow article for examples built around the `sim_` family.
