---
title: "Choosing and fitting disease progress models"
author: "Kaique S. Alves"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Choosing and fitting disease progress models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, message=FALSE, warning=FALSE}
library(epifitter)
library(dplyr)
library(ggplot2)
library(cowplot)
theme_set(cowplot::theme_half_open(font_size = 12))
```

## Overview

Disease progress models are simplified descriptions of epidemic development.
They are useful when the goal is to estimate biologically interpretable
features such as initial disease intensity, apparent infection rate, or an
upper asymptote. They should not be selected only because a curve looks smooth
or because a single fit statistic is largest.

This article explains the main model-fitting workflows in `epifitter`:

- `fit_lin()` for linearized model comparison;
- `fit_nlin()` for nonlinear fitting of two-parameter models;
- `fit_nlin2()` for nonlinear fitting when the asymptote `K` should be estimated.

Use `fit_lin()` for quick screening and teaching. Use nonlinear fitting when
you want to fit the model on the original disease intensity scale. Use
`fit_nlin2()` when the epidemic appears to plateau below 1 and estimating the
maximum disease intensity is scientifically meaningful.

## Compare models quickly with `fit_lin()`

Linearized models transform disease intensity before fitting a straight line.
This is fast and often useful for model comparison, but transformations also
change the error scale. Always inspect the fitted curves on the original
disease intensity scale.

```{r}
set.seed(1)
epi <- sim_logistic(N = 60, y0 = 0.01, dt = 5, r = 0.12, alpha = 0.2, n = 4)

fit_lin_out <- fit_lin(time = epi$time, y = epi$random_y)
knitr::kable(fit_lin_out$stats_all, digits = 4)
```

```{r fig.alt="Faceted plot comparing logistic and Gompertz fitted curves against observed disease intensity values."}
plot_fit(fit_lin_out, models = c("Logistic", "Gompertz"))
```

## Nonlinear fitting with starting values

`fit_nlin()` fits the same candidate models directly to the original disease
intensity values. Nonlinear optimization needs starting values, so poor starts
or weakly informative data can lead to non-convergence. When this happens,
inspect the curve, adjust `starting_par`, and consider whether the chosen model
is identifiable from the observed time points.

```{r eval=FALSE}
fit_nlin_out <- fit_nlin(
  time = epi$time,
  y = epi$random_y,
  starting_par = list(y0 = 0.01, r = 0.03)
)

fit_nlin_out$stats_all
```

## Weighted nonlinear fitting

By default, `fit_nlin()` and `fit_nlin2()` use ordinary nonlinear least
squares. This gives every observation the same weight and is consistent with
many traditional disease progress curve workflows. For disease incidence or
severity measured as a proportion, however, residual variability may change
across the epidemic. For example, binomial-like measurements often have
variance related to \(p(1 - p)\), where \(p\) is the fitted disease intensity.

`epifitter` therefore allows optional weighted nonlinear least squares. These
options are working variance assumptions, not a substitute for a full binomial
or beta likelihood. The default is unchanged:

```{r eval=FALSE}
fit_nlin(
  time = epi$time,
  y = epi$random_y,
  weight_method = "none"
)
```

To use a two-step binomial-like weighting approximation, use
`weight_method = "binomial"`. The model is first fitted without weights, fitted
values are used to construct weights, and the model is then refitted using
\(1 / (p(1 - p) + \epsilon)\).

```{r eval=FALSE}
fit_weighted <- fit_nlin(
  time = epi$time,
  y = epi$random_y,
  weight_method = "binomial",
  weight_eps = 0.01
)

fit_weighted$stats_all
```

Other built-in weighting rules are available for sensitivity analyses:

- `weight_method = "mean"` uses \(1 / (p + \epsilon)\), a mean-dependent variance approximation.
- `weight_method = "cv"` uses \(1 / (p^2 + \epsilon)\), corresponding to roughly constant coefficient of variation.
- `weight_method = "power"` uses \(1 / (|p|^{2\theta} + \epsilon)\), where `weight_power` controls \(\theta\).

These options follow the practical idea of modeling non-constant variance
without leaving the nonlinear least-squares framework. They are not universal
defaults. Use them as sensitivity analyses, and prefer the rule that best
matches how the disease intensity was measured.

```{r eval=FALSE}
fit_cv <- fit_nlin(
  time = epi$time,
  y = epi$random_y,
  weight_method = "cv"
)

fit_power <- fit_nlin(
  time = epi$time,
  y = epi$random_y,
  weight_method = "power",
  weight_power = 0.5
)
```

You can also provide your own positive weights. This is useful when weights
come from sample sizes, known measurement precision, or a variance model chosen
outside `epifitter`.

```{r eval=FALSE}
custom_weights <- rep(1, nrow(epi))

fit_custom <- fit_nlin(
  time = epi$time,
  y = epi$random_y,
  weights = custom_weights
)
```

For advanced use, `weights` can also be a function. The function receives a data
frame with `time`, `y`, `predicted`, and `model`, and must return one positive
weight per observation.

```{r eval=FALSE}
fit_custom_fun <- fit_nlin(
  time = epi$time,
  y = epi$random_y,
  weights = function(data) {
    1 / (data$predicted + 0.05)
  }
)
```

For grouped nonlinear fitting, use `weight_method` for the same strategy across
all curves, or `weights_col` when weights are stored in the data.

```{r eval=FALSE}
multi_weighted <- fit_multi(
  time_col = "time",
  intensity_col = "random_y",
  data = multi_epi,
  strata_cols = "curve",
  nlin = TRUE,
  weight_method = "binomial"
)
```

Weighted fits are assumptions, not automatic improvements. They are most useful
when residual diagnostics or the sampling design suggest heteroscedasticity.
Compare weighted and unweighted fits, inspect residuals, and report the chosen
weighting rule.

## Estimate `K` when the epidemic plateaus below 1

The two-parameter monomolecular, logistic, and Gompertz models assume a fixed
maximum disease intensity of 1. In real epidemics, disease may plateau below 1
because of host resistance, limited favorable conditions, management, sampling
scale, or a short observation window. In those cases, `fit_nlin2()` estimates
the upper asymptote `K`.

```{r eval=FALSE}
epi_partial <- epi %>%
  mutate(random_y = random_y * 0.8)

fit_k <- fit_nlin2(
  time = epi_partial$time,
  y = epi_partial$random_y,
  starting_par = list(y0 = 0.01, r = 0.03, K = 0.7)
)

fit_k$stats_all
```

## Grouped fitting with `fit_multi()`

Use `fit_multi()` when the same model-fitting workflow should be repeated for
many curves, such as treatment-by-block combinations, cultivars, sites, or
isolates. Keep grouping variables aligned with the experimental unit.

```{r}
epi1 <- sim_gompertz(N = 50, y0 = 0.001, dt = 5, r = 0.08, alpha = 0.2, n = 3)
epi2 <- sim_gompertz(N = 50, y0 = 0.002, dt = 5, r = 0.11, alpha = 0.2, n = 3)

multi_epi <- bind_rows(epi1, epi2, .id = "curve")

multi_fit <- fit_multi(
  time_col = "time",
  intensity_col = "random_y",
  data = multi_epi,
  strata_cols = "curve"
)

knitr::kable(head(multi_fit$Parameters), digits = 4)
knitr::kable(head(multi_fit$Data), digits = 4)
```

## Practical reporting

When reporting fitted disease progress models, include:

- the disease metric and scale, such as severity proportion or incidence proportion;
- the time scale and assessment schedule;
- the model family and whether `K` was fixed or estimated;
- whether the fit used ordinary or weighted nonlinear least squares;
- the weighting rule, if any;
- convergence issues or models that failed to fit;
- uncertainty intervals or sensitivity analyses when they inform interpretation.
