---
title: "Real-data workflow: powdery mildew in organic tomato"
author: "Kaique S. Alves"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Real-data workflow: powdery mildew in organic tomato}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, message=FALSE, warning=FALSE}
library(epifitter)
library(dplyr)
library(ggplot2)
library(cowplot)
theme_set(cowplot::theme_half_open(font_size = 12))
data("PowderyMildew")
```

## Overview

`PowderyMildew` is an experimental dataset included in `epifitter`. It contains disease progress curves for powdery mildew in organic tomato under different irrigation systems and soil moisture levels.

The data are derived from the study by Lage, Marouelli, and Cafe-Filho (2019) on powdery mildew management under different irrigation configurations in organic tomato.

The dataset is especially useful for demonstrating workflows with:

- a real epidemic dataset rather than simulated curves;
- repeated observations over time within experimental blocks;
- treatment comparisons using summary measures and fitted models.

This article is intended for users who want to move from raw experimental data
to interpretable epidemic summaries. The guiding principle is to keep the
experimental unit clear. Here, each block within an irrigation-by-moisture
treatment is repeatedly assessed over time.

## Load the data

Use the usual `data()` interface for packaged datasets.

```{r}
data("PowderyMildew")
knitr::kable(head(PowderyMildew, 12), digits = 4)
```

The dataset contains five variables:

- `irrigation_type` for the irrigation system;
- `moisture` for the soil moisture regime;
- `block` for the experimental block;
- `time` for the assessment date;
- `sev` for disease severity on a proportional scale.

## Inspect the experimental structure

```{r}
PowderyMildew %>%
  distinct(irrigation_type, moisture, block) %>%
  arrange(irrigation_type, moisture, block) %>%
  knitr::kable()
```

This is a repeated-measures dataset: each block was assessed across several time points within each irrigation-by-moisture combination.

For most treatment comparisons, the block-level curve is the unit that should
produce one summary value or one fitted model. Pooling all blocks together is
useful for visualization of treatment means, but it changes the inferential
target.

## Visualize treatment means over time

A useful first step is to summarize replicated blocks within each treatment and visualize the mean disease progress curve.

```{r}
pm_summary <- PowderyMildew %>%
  group_by(irrigation_type, moisture, time) %>%
  summarise(
    mean_sev = mean(sev),
    .groups = "drop"
  )

knitr::kable(head(pm_summary, 12), digits = 4)
```

```{r fig.alt="Line plot of mean powdery mildew severity over time for each irrigation treatment, faceted by soil moisture level."}
ggplot(pm_summary, aes(time, mean_sev, color = irrigation_type)) +
  geom_point(size = 1.8) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ moisture) +
  labs(
    title = "Mean powdery mildew progress by treatment",
    x = "Time",
    y = "Mean severity",
    color = "Irrigation"
  )
```

## Calculate AUDPC and AUDPS per experimental unit

Because `block` identifies repeated experimental units, a common analysis is to calculate one summary value per block within each treatment.

```{r}
pm_area <- PowderyMildew %>%
  group_by(irrigation_type, moisture, block) %>%
  summarise(
    audpc = AUDPC(time = time, y = sev, aggregate = "none"),
    audps = AUDPS(time = time, y = sev, aggregate = "none"),
    .groups = "drop"
  )

knitr::kable(pm_area, digits = 4)
```

This differs from pooling all rows together. If all blocks are passed at once, `AUDPC()` and `AUDPS()` assume that you want one treatment-level summary curve and aggregate repeated observations at each time point before calculating the area.

Statistically, these block-level values can later be analyzed with methods that
match the experimental design, such as models including treatment terms and
block structure. `epifitter` computes the epidemiological summaries; the
downstream treatment analysis should still respect randomization, blocking, and
replication.

To compare treatments visually, it is often useful to plot one point per block on top of a boxplot.

```{r fig.alt="Boxplots with jittered points showing the distribution of AUDPC and AUDPS values across irrigation treatments, faceted by moisture level and summary metric."}
pm_area_long <- pm_area %>%
  tidyr::pivot_longer(
    cols = c(audpc, audps),
    names_to = "summary_measure",
    values_to = "value"
  )

facet_aud <- if (requireNamespace("lemon", quietly = TRUE)) {
  lemon::facet_rep_grid(summary_measure ~ moisture, scales = "free_y")
} else {
  ggplot2::facet_grid(summary_measure ~ moisture, scales = "free_y")
}

ggplot(
  pm_area_long,
  aes(irrigation_type, value, fill = irrigation_type)
) +
  geom_boxplot(
    width = 0.65,
    outlier.shape = NA,
    alpha = 0.28,
    linewidth = 0.6
  ) +
  geom_jitter(
    aes(color = irrigation_type),
    width = 0.12,
    height = 0,
    size = 2.4,
    alpha = 0.85
  ) +
  facet_aud +
  labs(
    title = "Area-under-the-curve summaries by treatment",
    x = "Irrigation treatment",
    y = "Summary value",
    color = "Irrigation"
  ) +
  cowplot::theme_half_open(font_size = 12) +
  background_grid(major = "y", minor = "none") +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 20, hjust = 1),
    strip.background = element_rect(fill = "#e7f1f3", color = NA),
    strip.text = element_text(face = "bold")
  )
```

## Fit a model to one observed curve

To work with a single disease progress curve, first filter one treatment-by-block combination.

```{r}
single_curve <- PowderyMildew %>%
  filter(
    irrigation_type == "Drip",
    moisture == "High moisture",
    block == 1
  )

knitr::kable(single_curve, digits = 4)
```

```{r fig.alt="Scatter and line plot of a single observed powdery mildew disease progress curve for drip irrigation under high moisture in block 1."}
ggplot(single_curve, aes(time, sev)) +
  geom_point(size = 2, color = "#15616d") +
  geom_line(linewidth = 0.9, color = "#15616d") +
  labs(
    title = "Single observed disease progress curve",
    subtitle = "Drip irrigation, high moisture, block 1",
    x = "Time",
    y = "Severity"
  )
```

Now fit the candidate models to that curve.

The fitted parameters are descriptive summaries of the selected curve. For
example, the apparent infection rate is meaningful only in relation to the
model form, disease scale, and observed time interval.

```{r}
single_fit <- fit_lin(time = single_curve$time, y = single_curve$sev)
knitr::kable(single_fit$stats_all, digits = 4)
```

```{r fig.alt="Faceted plot comparing fitted disease progress curves for a single observed powdery mildew epidemic."}
plot_fit(single_fit, point_size = 2, line_size = 0.9)
```

## Fit models to many observed curves at once

`fit_multi()` is useful when each treatment-by-block combination should be fitted separately.

```{r}
pm_multi_fit <- fit_multi(
  time_col = "time",
  intensity_col = "sev",
  data = PowderyMildew,
  strata_cols = c("irrigation_type", "moisture", "block")
)

knitr::kable(head(pm_multi_fit$Parameters, 12), digits = 4)
```

This output keeps the original stratification variables, which makes it easier to compare fitted parameters among irrigation systems, moisture regimes, and blocks.

For formal treatment comparisons of fitted parameters, treat these outputs as
derived variables and analyze them using a model consistent with the trial
design. Avoid treating repeated assessments from the same block as independent
treatment replicates.

## Practical guidance

- Use `data("PowderyMildew")` to load the bundled dataset.
- Use grouped summaries by `block` when you want one AUDPC or AUDPS value per experimental unit.
- Use grouped means over time when the goal is a treatment-level visualization.
- Use `fit_lin()` for a quick model ranking on one observed curve.
- Use `fit_multi()` when you want to fit many treatment-block curves in a single workflow.

## Data source

The `PowderyMildew` dataset included in `epifitter` comes from:

Lage, D. A. C., Marouelli, W. A., and Cafe-Filho, A. C. (2019). Management of powdery mildew and behaviour of late blight under different irrigation configurations in organic tomato. *Crop Protection*, 125, 104886.
