---
title: "Simulating disease progress curves"
author: "Kaique S Alves"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating disease progress curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

In `epifitter`, the `sim_` family of functions complements the `fit_` functions by generating synthetic disease progress curve data from known model parameters. The same four population dynamics models supported for fitting can also be simulated.

These functions use `deSolve::ode()` (Soetaert, Petzoldt, and Setzer 2010) to solve the differential equation form of the epidemiological models.

## Hands On

### Packages

First, we need to load the packages we'll need for this tutorial.

```{r message=FALSE, warning=FALSE}
library(epifitter)
library(magrittr)
library(ggplot2)
library(cowplot)
theme_set(cowplot::theme_half_open(font_size = 12))
```

### The basics of the simulation

The `sim_` functions share the same core set of arguments. By default, only two are strictly required, while the others have sensible defaults.

-   `r`: apparent infection rate
-   `n`: number of replicates

When `n` is greater than one, replicated epidemics are produced and the `alpha` argument controls the level of noise (experimental variation). Together, these arguments generate `random_y` values that vary among replicates.

The other arguments are:

-   `N`: epidemic duration in time units
-   `dt`: time (fixed) in units between two assessments
-   `y0`: initial inoculum
-   `alpha`: noise parameter controlling replicate-level variation

#### Exponential

Let's simulate a curve resembling exponential growth.

```{r}
exp_model <- sim_exponential(
  N = 100,    # total time units 
  y0 = 0.01,  # initial inoculum
  dt = 10,    #  interval between assessments in time units
  r = 0.045,  #  apparent infection rate
  alpha = 0.2,# level of noise
  n = 7       # number of replicates
)
knitr::kable(head(exp_model), digits = 4)
```

A `data.frame` is produced with four columns:

-   `replicates`: replicate identifier
-   `time`: the assessment time
-   `y`: simulated disease intensity on a proportional scale
-   `random_y`: replicate-specific disease intensity after adding noise

Use [`ggplot2`](https://ggplot2.tidyverse.org/) to visualize the simulated curves.

```{r fig.alt="Plot of a simulated exponential disease progress curve with replicate observations over time."}
exp_plot = exp_model %>%
  ggplot(aes(time, y)) +
  geom_jitter(aes(time, random_y), size = 3,color = "gray", width = .1) +
  geom_line(size = 1) +
  theme_minimal_hgrid() +
  ylim(0,1)+
  labs(
    title = "Exponential",
    y = "Disease intensity",
    x = "Time"
  )
exp_plot
```

#### Monomolecular

The logic is exactly the same here.

```{r}
mono_model <- sim_monomolecular(
  N = 100,
  y0 = 0.01,
  dt = 5,
  r = 0.05,
  alpha = 0.2,
  n = 7
)
knitr::kable(head(mono_model), digits = 4)
```

```{r fig.alt="Plot of a simulated monomolecular disease progress curve with replicate observations over time."}
mono_plot = mono_model %>%
  ggplot(aes(time, y)) +
  geom_jitter(aes(time, random_y), size = 3, color = "gray", width = .1) +
  geom_line(size = 1) +
  theme_minimal_hgrid() +
  labs(
    title = "Monomolecular",
    y = "Disease intensity",
    x = "Time"
  )
mono_plot
```

#### The Logistic model

```{r}
logist_model <- sim_logistic(
  N = 100,
  y0 = 0.01,
  dt = 5,
  r = 0.1,
  alpha = 0.2,
  n = 7
)
knitr::kable(head(logist_model), digits = 4)
```

```{r fig.alt="Plot of a simulated logistic disease progress curve with replicate observations over time."}
logist_plot = logist_model %>%
  ggplot(aes(time, y)) +
  geom_jitter(aes(time, random_y), size = 3,color = "gray", width = .1) +
  geom_line(size = 1) +
  theme_minimal_hgrid() +
  labs(
    title = "Logistic",
    y = "Disease intensity",
    x = "Time"
  )
logist_plot
```

#### Gompertz

```{r}
gomp_model <- sim_gompertz(
  N = 100,
  y0 = 0.01,
  dt = 5,
  r = 0.07,
  alpha = 0.2,
  n = 7
)
knitr::kable(head(gomp_model), digits = 4)
```

```{r fig.alt="Plot of a simulated Gompertz disease progress curve with replicate observations over time."}
gomp_plot = gomp_model %>%
  ggplot(aes(time, y)) +
  geom_jitter(aes(time, random_y), size = 3,color = "gray", width = .1) +
  geom_line(size = 1) +
  theme_minimal_hgrid() +
  labs(
    title = "Gompertz",
    y = "Disease intensity",
    x = "Time"
  )
gomp_plot
```

### Combo

Use `plot_grid()` from [`cowplot`](https://wilkelab.org/cowplot/index.html) to combine the simulated curves in a grid.

```{r fig.height=6, fig.width=8, fig.alt="Grid combining the exponential, monomolecular, logistic, and Gompertz simulated disease progress curves."}
plot_grid(exp_plot,
          mono_plot,
          logist_plot,
          gomp_plot)
```

## References

Karline Soetaert, Thomas Petzoldt, R. Woodrow Setzer (2010). Solving Differential Equations in R: Package deSolve. Journal of Statistical Software, 33(9), 1--25. DOI: [10.18637/jss.v033.i09](http://dx.doi.org/10.18637/jss.v033.i09)
