---
title: "Simulation workflows for disease progress curves"
author: "Kaique S. Alves"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Simulation workflows for disease progress curves}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, message=FALSE, warning=FALSE}
library(epifitter)
library(ggplot2)
library(cowplot)
theme_set(cowplot::theme_half_open(font_size = 12))
```

## Overview

The `sim_` family creates synthetic disease progress curves that match the same model families used by the fitting functions.

Simulations are useful for teaching, testing analysis workflows, checking
whether a model can recover known parameters, and generating reproducible
examples. They are not a substitute for biological validation. Choose parameter
values that make sense for the host-pathogen system, disease metric, and time
scale.

## Simulate four canonical curve shapes

The four model families represent different epidemic shapes. Exponential
growth is most appropriate for early unconstrained increase, while
monomolecular, logistic, and Gompertz curves include different forms of
deceleration as disease approaches an upper limit.

```{r}
exp_model <- sim_exponential(N = 100, y0 = 0.01, dt = 10, r = 0.045, alpha = 0.2, n = 5)
mono_model <- sim_monomolecular(N = 100, y0 = 0.01, dt = 5, r = 0.05, alpha = 0.2, n = 5)
log_model <- sim_logistic(N = 100, y0 = 0.01, dt = 5, r = 0.10, alpha = 0.2, n = 5)
gomp_model <- sim_gompertz(N = 100, y0 = 0.01, dt = 5, r = 0.07, alpha = 0.2, n = 5)
```

```{r}
exp_plot <- ggplot(exp_model, aes(time, y)) +
  geom_jitter(aes(y = random_y), width = 0.1, color = "#6c757d") +
  geom_line(color = "#b56576", linewidth = 0.8) +
  labs(title = "Exponential")

mono_plot <- ggplot(mono_model, aes(time, y)) +
  geom_jitter(aes(y = random_y), width = 0.1, color = "#6c757d") +
  geom_line(color = "#588157", linewidth = 0.8) +
  labs(title = "Monomolecular")

log_plot <- ggplot(log_model, aes(time, y)) +
  geom_jitter(aes(y = random_y), width = 0.1, color = "#6c757d") +
  geom_line(color = "#355070", linewidth = 0.8) +
  labs(title = "Logistic")

gomp_plot <- ggplot(gomp_model, aes(time, y)) +
  geom_jitter(aes(y = random_y), width = 0.1, color = "#6c757d") +
  geom_line(color = "#8d5a97", linewidth = 0.8) +
  labs(title = "Gompertz")
```

```{r fig.height=6, fig.width=8, fig.alt="Grid of four simulated disease progress curves showing exponential, monomolecular, logistic, and Gompertz shapes."}
plot_grid(exp_plot, mono_plot, log_plot, gomp_plot, ncol = 2)
```

## Send simulated data into the fitting pipeline

```{r}
fit_from_sim <- fit_lin(time = log_model$time, y = log_model$random_y)
fit_from_sim$stats_all
```

When using simulations for method checking, compare the fitted model with the
known data-generating model. When using simulations for teaching, show both the
underlying curve (`y`) and the noisy observations (`random_y`) so users can
distinguish the epidemic process from the observation process.
